Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-03-03 docs/manuscript/manuscript.tex

We preprocessed the data, and we recognized that there were three different surveys collected by the DHS program via excel. They are DHS, MIS, and AIDS Indicator Survey (AIS). After, we used R to develop an algorithm to find the most recent surveys and showed results for 78 countries.  We also simplified the variables titles into four character acronyms to recognize the different 28 variables quickly. We used the household data only. The STAT-compiler collected the aggregated data by the 78 countries

We preserved 28 variables from 73 countries where less than 50\% of values were missing and the population survey data were removed because they were similar to the household surveys. Ultimately, the simplified data frame was 78 countries by 28 variables. After the simplification, two separate tables are distinguished within the 28 variables. One focused on the water accessibility variables and the other focused on the explanatory variables \citep{price2019difference}. From the remaining variables, we selected 17 which related to water accessibility.  These included: types of water source (piped into dwelling, piped into yard/plot,
 public tap/standpipe, tube well/borehole, protected well, protected spring, rainwater, unprotected well, unprotected spring, cart with small tank/drum, tanker truck, bottled water, other water source, surface water), time to water source (less than 30 minutes, more than 30 minutes), and premise to water source (water on the premises) \autoref{tab:summary-wa}. Remaining variables related to wealth and transportation were used as an explanatory variables in the model \autoref{tab:summary-var}.

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
 \texttt{phom} & Piped into dwelling & DHS Program Surveys
 \\\midrule
 \texttt{pipy} & Water piped into yard/plot
& DHS Program Surveys
\\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & DHS Program Surveys
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pwel} & Protected well
& DHS Program Surveys
\\\midrule
\texttt{pspr} & Protected spring
& DHS Program Surveys
\\\midrule

\texttt{rain} & Rainwater
& DHS Program Surveys
\\\midrule
\texttt{uwel} & Unprotected well water
& DHS Program Surveys
\\\midrule
\texttt{uspr} & Unprotected spring
& DHS Program Surveys
\\\midrule

 \texttt{truc} & Tanker truck & DHS Program Surveys
\\\midrule
 \texttt{ctan} & Cart with tank & DHS Program Surveys
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john (DHS Program Surveys).
\\\midrule
 \texttt{othw} & Other water source & DHS Program Surveys
\\\midrule
\texttt{surw} & Surface water
& DHS Program Surveys
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
& DHS Program Surveys
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & DHS Program Surveys
\\\midrule
 \texttt{watp} & Water on the premises
 & DHS Program Surveys
 \\\bottomrule
  \end{tabular}
  \caption{Summary of water accessibility variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-wa}
\end{table}

@@

Across 78 countries in this study, the data provides 28 quantitative variables and three descriptive variables such as the country names, survey type, and survey year. The majority of the data is from 2010 - 2020 as shown in \autoref{fig:year}. Therefore, we can analyze the global condition of water accessibility by utilizing the most up-to-date surveys. Most of the variables are related to water accessibility by their locations, time, and water resources. This also allows us to find water accessibility predictor variables and new clustering graphs.

@@

\subsection{Cluster analysis}
We selected 22 variables that describe water accessibility of 78 countries around the world \citep{price2019difference}. Next, we used clustering, an unsupervised machine learning method to identify and group water accessibility variables in a large data set to understand and make inferences using R. Clustering ultimately can continue to assist in discovering the water accessibility predictor variables.

>>>>

We preprocessed the data, and we recognized that there were three different surveys collected by the DHS program via excel. They are DHS, MIS, and AIDS Indicator Survey (AIS). After, we used R to develop an algorithm to find the most recent surveys and showed results for 73 countries.  We also simplified the variables titles into four character acronyms to recognize the different 28 variables quickly. We used the household data only. The STAT-compiler collected the aggregated data by the 73 countries

We preserved 28 variables from 73 countries where less than 50\% of values were missing and the population survey data were removed because they were similar to the household surveys. Ultimately, the simplified data frame was 73 countries by 28 variables. After the simplification, two separate tables are distinguished within the 28 variables. One focused on the water accessibility variables and the other focused on the explanatory variables \citep{price2019difference}. From the remaining variables, we selected 17 which related to water accessibility.  These included: types of water source (piped into dwelling, piped into yard/plot,
 public tap/standpipe, tube well/borehole, protected well, protected spring, rainwater, unprotected well, unprotected spring, cart with small tank/drum, tanker truck, bottled water, other water source, surface water), time to water source (less than 30 minutes, more than 30 minutes), and premise to water source (water on the premises) \autoref{tab:summary-wa}. Remaining variables related to wealth and transportation were used as an explanatory variables in the model \autoref{tab:summary-var}.

\begin{table}[h!]
  \centering
  \begin{tabular}{p{1in} p{3in} p{2in}}\toprule
    \bf Variable & \bf Description & \bf Notes \\\midrule
 \texttt{phom} & Piped into dwelling & 
 \\\midrule
 \texttt{pipy} & Water piped into yard/plot
& 
\\\midrule
    \texttt{ptap} &  Public tap/standpipe
 & 
 \\\midrule
 \texttt{bore} & Tube well/borehole & 
\\\midrule
\texttt{pwel} & Protected well
& 
\\\midrule
\texttt{pspr} & Protected spring
& 
\\\midrule

\texttt{rain} & Rainwater
& 
\\\midrule
\texttt{uwel} & Unprotected well water
&
\\\midrule
\texttt{uspr} & Unprotected spring
& 
\\\midrule

 \texttt{truc} & Tanker truck & 
\\\midrule
 \texttt{ctan} & Cart with tank & 
\\\midrule
 \texttt{bott} & Bottled water & Drinking water is bottled water/demi john.
\\\midrule
 \texttt{othw} & Other water source & 
\\\midrule
\texttt{surw} & Surface water
& 
\\\midrule
\texttt{tles} & Water 30 minutes or less away round trip
& 
\\\midrule 
\texttt{tmor} & Water more than 30 minutes away round trip & DHS Program Surveys
\\\midrule
 \texttt{watp} & Water on the premises
 & 
 \\\bottomrule
  \end{tabular}
  \caption{Summary of water accessibility variables (National Household usage Averages measured as percentage for Main Drinking Source)}
  \label{tab:summary-wa}
\end{table}

@@

Across 73 countries in this study, the data provides 28 quantitative variables and three descriptive variables such as the country names, survey type, and survey year. The majority of the data is from 2010 - 2020 as shown in \autoref{fig:year}. Therefore, we can analyze the global condition of water accessibility by utilizing the most up-to-date surveys. Most of the variables are related to water accessibility by their locations, time, and water resources. This also allows us to find water accessibility predictor variables and new clustering graphs.

@@

\subsection{Cluster analysis}
We selected 17 variables that describe water accessibility of 73 countries around the world \citep{price2019difference}. Next, we used clustering, an unsupervised machine learning method to identify and group water accessibility variables in a large data set to understand and make inferences using R. Clustering ultimately can continue to assist in discovering the water accessibility predictor variables.
